---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo=TRUE
)
library(rsimpop)
```
---
date: `r format(Sys.Date(), "%d/%m/%Y")`
---


# rsimpop

<!-- badges: start -->
<!-- badges: end -->


## Installation

You can install rsimpop like so:

``` r
devtools::install_github("NickWilliamsSanger/rsimpop")
```

Or clone the repository and install using the downloaded package tar file

``` r
install.packages("somedirectory/rsimpop/rsimpop_2.2.0.tar.gz",repos=NULL)
```

## rsimpop
This package implements Gillespie Algorithm implementation of the Birth-Death model that allows the simultaneous simulation of multiple cellular compartments each with their own time varying-target population size and sub-compartments with differential fitness (driver compartments).


## Simulate from Zygote for 1 year and subsample tree
```{r zygote,results=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}

##Initialise with seed (R and rsimpop separately)
SEED=37774323
initSimPop(SEED,bForce = TRUE)
##Setup a single compartment with a target pop of 50K
cfg=getDefaultConfig(target_pop_size  = 5e4,ndriver = 1,basefit = 0.2,rate = 0.1)
print(cfg)
##Simulate for 2years..
sp=sim_pop(NULL,params=list(n_sim_days=365*2,b_stop_at_pop_size=1),cfg=cfg)
##Look at the population size trajectory
plot(sp)
##Subsample tree
sampledtree1=get_subsampled_tree(sp,100)
print(sampledtree1)
plot_tree(sampledtree1,cex.label = 0.5)
title("Sampled Early Development Tree: Division Tree")
```

Notice how the sampled tree has 101 tips rather than the specified 100.  This is because the simulator always maintains an inactive outgroup (here s1). A group is rendered inactive by specifying a negative "rate" in the cfg$compartment dataframe.  The tree branch lengths are now given in terms of the number of self renewal divisions.  This allows the user to flexibly apply their own mutation acquisition model:

```{r mutacq}
get_elapsed_time_tree
sampledtree1m=get_elapsed_time_tree(sampledtree1,mutrateperdivision=1,backgroundrate=15/365)
```

```{r events, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
t1=plot_tree(sampledtree1m,cex.label = 0.5);title("Sampled Early Development Tree: Mutation Tree")
node_labels(t1,cex=0.8)
```
The blue numbers are represent the ID of the child node associated with each branch. This "node" ID together with a timestamp is used to locate events on the tree - see, the *events* data.frame , that is maintained and updated by the simulator:   

```{r}
print(sampledtree1m$events)
```

Notice how here the events dataframe specifies the compartment for the outgroup and the rest of the tree.

We can introduce another cell compartment as follows:

```{r newcompartment,message=TRUE}
cfg=sampledtree1$cfg
cfg=addCellCompartment(cfg,population = 5e4,rate=1/50,ndriver=1,descr="MyTissue",basefit = 0.3)
cfg$compartment$rate[2]=1/120  ## change the rate of compartment 1
sampledtree1a=addDifferentiationEvents(sampledtree1,cfg,2,nEvent=10)
print(sampledtree1a$events)
```

The occurence of such events for such sub-sampled trees can be visualised using the built in function *plot_tree_events*

```{r plotc2, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
plot_tree_events(sampledtree1a)
```

Recall that the config has already been updated with the target population sizes and division rates.  The next phase can therefore be simulated:

```{r simagain,message=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
sp2=sim_pop(sampledtree1a,params=list(n_sim_days=365*10),cfg=sampledtree1a$cfg)
plot(sp2)
sampledtree2=get_subsampled_tree(sp2,100)
plot_tree_events(sampledtree2,cex.label = 0.5)
```

## Wrapper function for neutral simulation

```{r rneutw,dev.args=list(pointsize=8),fig.height=5,fig.width=10}
run_neutral_sim
testing=run_neutral_sim(0.1,0.5/365,target_pop_size = 1e3,nyears=10)
plot(testing)
st=get_subsampled_tree(testing,100)
print(st$cfg$info)
plot_tree_events(st,cex.label = 0);title("Phylogeny of 100 cells sampled from small population")
```

## Selection based simulation

Here we are interested in the simple situation of one cellular compartment with multiple sub-compartments.

```{r selection, echo=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
#run_selection_sim
selsim=run_selection_sim(0.05,1/(2*190),target_pop_size = 5e4,nyears = 50,fitness=0.3)
print(selsim$cfg$info)
```

Plot example sampled trees using various brach length models (or scaling):

```{r seltree, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
seltree100=get_subsampled_tree(selsim,100)
print(seltree100$cfg$info)
plot_tree_events(seltree100,cex.label = 0);title("Selection Based Tree: Branch Length=#Self Renewal Divisions")
seltree100rt=get_elapsed_time_tree(seltree100)
tree=plot_tree_events(seltree100rt,cex.label = 0);title("Selection Based Tree: Branch Length=#Real Time")
mp=5
seltree100m=get_elapsed_time_tree(seltree100,mutrateperdivision=mp,backgroundrate=(20-(365/190)*mp)/365)
plot_tree_events(seltree100m,cex.label = 0.5);title("Selection Based Tree: Branch Length=#Mutations")
seltree100m2=get_elapsed_time_tree(seltree100,mutrateperdivision=20*(190/365),backgroundrate=0)
plot_tree_events(seltree100m2,cex.label = 0.5);title("Selection Based Tree: Branch Length=#Mutations v2")
```

## Transient selection

```{r selectiont,error=TRUE,echo=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
#run_transient_selection
tselsim=run_transient_selection(0.05,1/(2*190),target_pop_size = 5e4,nyears_driver_acquisition=15,
                                  nyears_transient_end=30,
                                  nyears=50,
                                  fitness=0.5)
tseltree200=get_subsampled_tree(tselsim,200)
plot_tree_events(get_elapsed_time_tree(tseltree200),cex.label=0)
```


## Neutral simulation with a trajectory

Create a trajectory dataframe with 3 columns (ts,target_pop_size,division_rate) and simulate using the run_neutral_trajectory wrapper function. Note that timestamps and rates are expressed in units of days and expected divisions per day respectively.

```{r neutwithtraj, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
trajectory=data.frame(ts=365*(1:80),target_pop_size=5e4+100*(1:80),division_rate=1/(2*190))
trajectory$target_pop_size[5:10]=2*trajectory$target_pop_size[5:10]
trajectory$target_pop_size[11:15]=0.2*trajectory$target_pop_size[11:15]
print(head(trajectory))
sp=run_neutral_trajectory(NULL,0.5,trajectory)
plot(sp,xlim=c(0,100))
lines(trajectory$ts/365,trajectory$target_pop_size,col="red")
legend("topright",c("Target","Actual"),col=c("red","black"),lwd=1)
st=get_subsampled_tree(sp,100)
plot_tree(get_elapsed_time_tree(st),cex.label = 0)
```
## Neutral simulation with a trajectory and a non-zero death rate
By default during the growth phase of the simulation the population is a pure birth process until the target population is hit at which point the aggregate death rate is set to match the aggregate birth rate. A death rate (less than birth rate) can also be specified in the trajectory and the population will then stochastically grow exponentially towards the target following a birth-death process.

```{r neutwithtrajdr, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
trajectory=data.frame(ts=c(70,400),target_pop_size=c(1e4,1e4),division_rate=c(0.06,0.06),death_rate=c(0.05,0.05))
print(head(trajectory))
sp=run_neutral_trajectory(NULL,initial_division_rate = 0.1,trajectory = trajectory)
plot(sp)
lines(trajectory$ts,trajectory$target_pop_size,col="red")
legend("topright",c("Target","Actual"),col=c("red","black"),lwd=1)
abline(v=trajectory$ts,lwd=2,col="blue")
st=get_subsampled_tree(sp,200)
plot_tree(get_elapsed_time_tree(st),cex.label = 0)
```

## Multiple drivers

Multiple drivers can be generated at a specified rate so the waiting time between events is exponentially distributed.

Firstly the user need to create a function that draws a selection coefficient from a distibution. The simulator can comfortably accommodate over 100 drivers per year but nonetheless it is advised that a selection coefficient threshold be specified (say 0.05) and that driver incidence be made correspondingly rarer. Note that at least 200,000 fitness coefficent values need to provided to run_driver_process_sim.

### Selection coefficient specification

```{r}
##Function to generate exponential distribution based fitness
require("truncdist")
genExpFitness=function(fitness_threshold,rate){
    function() rtrunc(n=1,a=fitness_threshold, b=Inf,"exp",rate=rate)
}
fitnessExpFn=genExpFitness(fitness_threshold=0.08,rate=40)
hist(sapply(1:10000,function(x) exp(fitnessExpFn())-1),breaks=seq(0,100,0.01),xlim=c(0,1),xlab="Selection Coefficient Per Year",main="Sampled Selection Cofficent Distribution")
```

### Run the multiple driver simulation
Introduce one driver per year 
```{r,message=FALSE,results=FALSE}
dps=run_driver_process_sim(simpop=NULL,initial_division_rate = 0.1,final_division_rate = 1/(2*190),target_pop_size = 1e5,nyears = 80,fitness=fitnessExpFn,drivers_per_cell_per_day = 1/(365*1e5))
```

Look at the final per driver counts

```{r}
print(dps$cfg$info %>% filter(population>0))
```

Plot an example sampled tree

```{r, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
dpst=get_subsampled_tree(dps,200)
dpst=get_elapsed_time_tree(dpst)
plot_tree_events(dpst)
dpst=get_elapsed_time_tree(dpst,mutrateperdivision = 1,backgroundrate = 19/365)
plot_tree_events(dpst,cex.label = 0,fmode=1)
```

Continue simulating the same individual until the age of 90

```{r}
#dps90=continue_driver_process_sim(dps,25,fitness_gen=fitnessExpFn(),drivers_per_year = 0.5)  ##continue_driver_process_sim(dps,90,fitnessGen =fitnessExpFn(),)
dps90=run_driver_process_sim(simpop=dps,initial_division_rate = 0.1,final_division_rate = 1/(2*190),target_pop_size = 1e5,nyears = 90,fitness=fitnessExpFn,drivers_per_cell_per_day = 1/(365*1e5))
```

Note that as of version 2.0.0 driver ids are not reused so they are now preserved between runs.
```{r, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
dpst90=get_subsampled_tree(dps90,200)
dpst90=get_elapsed_time_tree(dpst90)
plot_tree_events(dpst90,cex.label = 0)
```

If desired the colouring can be kept consistent across 2 plots from different times (experimental) using the events.keep argument:

```{r, dev.args=list(pointsize=8),fig.height=8,fig.width=10}
par(mfcol=c(2,1))
zz=plot_tree_events(dpst,fmode=1,cex.label = 0)
yy=plot_tree_events(dpst90,fmode=1,events.keep = zz$events,cex.label = 0)
```

## Multiple drivers with Trajectory

```{r,dev.args=list(pointsize=8),fig.height=6,fig.width=8}
trajectory=data.frame(ts=365*(1:80),target_pop_size=5e4+100*(1:80),division_rate=1/(2*190),compartment=1)
trajectory$target_pop_size[15:20]=2*trajectory$target_pop_size[15:20]
trajectory$target_pop_size[21:25]=0.2*trajectory$target_pop_size[21:25]
dps=run_driver_process_sim(initial_division_rate=0.1,user_trajectory=trajectory, target_pop_size = 1e5,nyears = 80,
                           fitnessGen=fitnessExpFn,drivers_per_cell_per_day = 1/(365*1e5))

plot(dps,xlim=c(0,100))
lines(trajectory$ts/365,trajectory$target_pop_size,col="red")
legend("topright",c("Target","Actual"),col=c("red","black"),lwd=1)
dpst=get_subsampled_tree(dps,200)
dpst=get_elapsed_time_tree(dpst)
plot_tree_events(dpst,cex.label = 0)
```

## Modelling continuous differentiation/migration between cellular compartments

```{r, dev.args=list(pointsize=8),fig.height=6,fig.width=10}
cfg=getDefaultConfig(target_pop_size  = 1e4,rate = 0.1)
print(cfg)
cfg$compartment$desc[2]="EMB"
##Simulate rapid growth before HSC specification
sp=sim_pop(NULL,params=list(n_sim_days=365*1,b_stop_at_pop_size=1),cfg=cfg)
##We say the HSC pool is started with 50 cells
HSC_INIT_POP=50
sp=get_subsampled_tree(sp,HSC_INIT_POP)
cfg=sp$cfg
cfg=addCellCompartment(cfg,population = 5e4,rate=0.02,descr="HSC")
cfg=addCellCompartment(cfg,population = 5e5,rate=0.04,descr="MPP")
cfg$compartment$rate[2]=c(-1)  ## change the rate of compartment 1
cfg$compartment$popsize[2]=c(0) ## switch off EMB compartment
sp=addDifferentiationEvents(sp,cfg,2,nEvent=HSC_INIT_POP,currentCompartment = 1)
## Specify the HSC->MPP migrations (just symmetric at a low rate)
sp$cfg$migrations=data.frame(c1=c(2),## source: HSC
                             c2=c(3),## sink: MPP
                             arate=c(0), ## asymmetric diff rate
                             srate=c(0.001) ## symmetric diff rate
                             )
sp=sim_pop(sp,params=list(n_sim_days=2*365),cfg=sp$cfg)
plot_compartment_trajectory(sp)
plot_compartment_trajectory(sp)+scale_y_log10()
st=get_subsampled_tree_fixed_compartmentcount(sp,compartments = c(2,3),counts=c(150,150))
plot_tree_events(get_elapsed_time_tree(st,0,50/365),cex.label=0,use.desc=TRUE)
```

